Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos, Crete

### 2021-02-19 analysis/paper/Family_Fortunes_Article.Rmd
Edits

---
title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
author:
  - Dominic Pollard
  - Institute of Archaeology, University College London, 31-34 Gordon Square, Bloomsbury, London WC1H 0PY, United Kingdom
  - dominic.pollard.16@ucl.ac.uk
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-anthropological-archaeology.csl" # Insert path for the bib-style

@@

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
library(ggpubr)
library(ineq)
library(reshape)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(jpeg)
library(ggrepel)

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].

@@

```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T. Whitelaw)."}

@@

Survey aside, our most abundant evidence comes from the mortuary record (see Figure \@ref(fig:map)). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with burials known from the Kephala ridge [@Coldstream1963: 38; @Hogarth1899: 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968: 133–46], the cave of Mavro Spelio [@Antoniadis2012: 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997 @Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed that they provide little scope for systematic quantitative analysis.

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. TAken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.

@@

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably the southernmost burials of the KNC. Unfortunately, all had been thoroughly looted at the time of excavation and are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].

The tombs themselves fall largely into three types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-48]. The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted solely to child burials [@Antoniadis2012: 69].

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA. For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123]. Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and accorded exceptional treatment in death [@Catling1995: 127–8].

@@

## 2.6 Number Crunching: Quantitative Analyses of the Cemeteries

@@

```{r allpithoi,fig.dim = c(8, 3), dpi=600, fig.align='center', fig.cap="(Left) Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods. (Right) Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time."}

@@

non_aorist_fig<-ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram(fill="gray40")+
  theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11))+
          labs(x="Date (Years BC)", y="Vessel Count")+
                theme(legend.position="none")+
                  scale_y_continuous(expand=c(0,0),limits=c(0,610))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

aorist_all_tombs<-colSums(tombs[158:207])
aorist_all_tombs<-data.frame(aorist_all_tombs)

@@

aorist_fig<-ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
  geom_col(fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11))+
          labs(x="Date (Years BC)", y="Aoristic Sum")+
               theme(legend.position="none")+
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

ggarrange(non_aorist_fig,aorist_fig,ncol=2,common.legend=FALSE)

@@

```{r aoristbycem, dpi=600, fig.align="center", fig.cap="Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,3.5)}
KNC<-subset(tombs,tombs$cem_condensed=="KNC")

aorist_KNC_pithoi <- colSums(KNC[208:257])
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)

@@

aorist_FOR_pithoi <-colSums(FOR[208:257])
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)

@@

aorist_all_pithoi <- colSums(tombs[208:257])
aorist_all_pithoi<-data.frame(aorist_all_pithoi)

@@

aorist_pithoi_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

aorist_KNC <- colSums(KNC[158:207])
aorist_KNC<-data.frame(aorist_KNC)

@@

aorist_FOR <-colSums(FOR[158:207])
aorist_FOR<-data.frame(aorist_FOR)

@@

aorist_pots_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
```

And, despite differences in scale and timing, it is immediately apparent from Figure \@ref(fig:aoristbycem) that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods. In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th. Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity. We might assume that these trends would be mirrored in the construction of new tombs, but this proves not to be the case.

```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations. In both cases, tombs lacking any securely dateable pottery are excluded. Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising."}

@@

new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position=position_dodge2(preserve="single"))+theme_classic()+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
            labs(x="Period", y="Count of New Tombs")+
              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                 scale_y_continuous(expand=c(0,0),limits=c(0,30))

new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position=position_dodge2(preserve="single"))+
  theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
            labs(x="Period", y="Aoristic Sum of New Tombs")+
               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                 scale_y_continuous(expand=c(0,0),limits=c(0,30))

ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,common.legend=TRUE, legend="bottom")

@@

KNC_tombs_in_use<-colSums(KNC[6:15]!="")
KNC_tombs_in_use<-data.frame(KNC_tombs_in_use)
colnames(KNC_tombs_in_use)<-c("no_tombs")
setDT(KNC_tombs_in_use, keep.rownames = "period")
KNC_tombs_in_use$period<-factor(KNC_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
KNC_tombs_in_use$cemetery<-"KNC"

FOR_tombs_in_use<-colSums(FOR[6:15]!="")
FOR_tombs_in_use<-data.frame(FOR_tombs_in_use)
colnames(FOR_tombs_in_use)<-c("no_tombs")
setDT(FOR_tombs_in_use, keep.rownames = "period")
FOR_tombs_in_use$period<-factor(FOR_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
FOR_tombs_in_use$cemetery<-"FOR"

@@

ggplot(tombs_in_use_cem,aes(period,no_tombs, fill=cemetery))+geom_col(position="dodge")+theme_classic()+
       theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
            labs(x="Period", y="Tombs in Use")+
              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                 scale_y_continuous(expand=c(0,0),limits=c(0,45))
```

@@

```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="Estimates of the number of tombs in use through time across both cemeteries. The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
tombs_11th<-subset(tombs,tombs$century_built=="11th")
tombs_10th<-subset(tombs,tombs$century_built=="10th")
tombs_9th<-subset(tombs,tombs$century_built=="9th")
tombs_8th<-subset(tombs,tombs$century_built=="8th")
tombs_7th<-subset(tombs,tombs$century_built=="7th")

@@

ggplot(data=tomb_use,aes(period,no_tombs))+geom_col(alpha=0.2)+
            geom_col(data=tomb_use,aes(period,no_tombs,fill=tomb_century),position="dodge")+
              theme_bw()+theme(text=element_text(family="serif", size=12))+
                 labs(x="Period", y="Estimated No. of Tombs in Use")+
                  scale_fill_brewer(name="Century Built",palette = "Set2")+ 
                    scale_y_continuous(expand=c(0,0),limits=c(0,60))

@@

```{r potsbydateandtombdate,  fig.dim=c(3,7),out.width="0.5%", dpi=600, fig.cap="Counts of vessels from all tombs in both cemeteries, arranged, above, by the century to which they are assigned based on their ceramic period and, below, by the centuries in which the tombs they come from were built."}

@@

>>>>

---
title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos, Crete"
author:
  - Dominic Pollard
  - Institute of Archaeology, University College London, 31-34 Gordon Square, Bloomsbury, London WC1H 0PY, United Kingdom
  - dominic.pollard.16@ucl.ac.uk
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-anthropological-archaeology.csl" # Insert path for the bib-style

@@

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
library(ggpubr)
library(ineq)
library(reshape2)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(jpeg)
library(ggrepel)
library(forcats)
library(factoextra)
library(cluster)

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric (G) date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].

@@

```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA tombs (crosses) and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project (circles), giving an impression of the contemporary urban extent (underlying image credit: T. Whitelaw)."}

@@

Survey aside, our most abundant evidence comes from the mortuary record (see Figure \@ref(fig:map)). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with burials known from the Kephala ridge [@Coldstream1963: 38; @Hogarth1899: 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968: 133–46], the cave of Mavro Spelio [@Antoniadis2012: 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997;@Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed that they provide little scope for systematic quantitative analysis.

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries (Figure \@ref(fig:map2)), which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. Taken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.

@@

```{r map2, fig.align= "centre", fig.cap="Closer view of the Knossos North and Fortetsa cemeteries. Tombs were digitised and georeferenced from the original publications, Coldstream and Catling (1996) and Brock (1957)."}

knitr::include_graphics("../figures/Knossos Tombs.jpg")
```

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably the southernmost burials of the KNC. Unfortunately, all had been thoroughly looted at the time of excavation and are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site (KNC Main and NW in Figure \@ref(fig:map2)), named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].

The tombs themselves fall largely into three types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-48]. The vast majority of burials are cremations deposited in pithoi or other urns, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted solely to child burials [@Antoniadis2012: 69].

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA. For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123]. Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and so were accorded exceptional treatment in death [@Catling1995: 127–8].

@@

## 2.6 Quantitative Analyses of the Cemeteries

@@

```{r allpithoi,fig.dim = c(7, 7), dpi=600, fig.align='center', fig.cap="(A) Counts of all vessels per decade across both cemeteries. Vessel dates have here been based on the mean of the absolute date range corresponding to their respective ceramic periods. (B) The same but with aoristic sums replacing raw counts. This represents a probabilistically weighted visualisation of depositional activity through time."}

@@

rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
                    xend =    c(970,920,875,840,810,790,745,710,670,630,600), 
                    xmid =   c(1035,945,897.5,857.5,825,800,767.5,727.5,690,650,615),
                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))

non_aorist_fig<-ggplot()+geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 590, ymax = 620), fill=NA, colour="black",alpha = 0.4) +
  geom_histogram(data=pots, aes(ave_date,fill="Pots"),fill="gray40",binwidth = 10)+
  geom_text(data=rects,aes(xmid, 607,label=col),family="serif",size=3)+
  theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11))+
          labs(x="Date (Years BC)", y="Vessel Count")+
                theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
                  scale_y_continuous(expand=c(0,0),limits=c(0,620))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

aorist_all_tombs<-colSums(tombs[160:209])
aorist_all_tombs<-data.frame(aorist_all_tombs)

@@

aorist_fig<-ggplot()+
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 214, ymax = 225), fill=NA, colour="black",alpha = 0.4) +
  geom_text(data=rects,aes(xmid, 220,label=col),family="serif",size=3)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic
  Sum"),fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11))+
          labs(x="Date (Years BC)", y="Aoristic Sum")+
              theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

ggarrange(non_aorist_fig,aorist_fig,nrow=2,labels=c("A","B"),align="v",common.legend=FALSE)

@@

```{r aoristbycem, dpi=600, fig.align="center", fig.cap="(A) Aoristic sums of pithoi, as a proxy for cremation burials, through time. (B) Aoristic sums of all vessels through time.", fig.dim=c(7,7.5)}
KNC<-subset(tombs,tombs$cem_condensed=="KNC")

aorist_KNC_pithoi <- colSums(KNC[210:259])
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)

@@

aorist_FOR_pithoi <-colSums(FOR[210:259])
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)

@@

aorist_all_pithoi <- colSums(tombs[210:259])
aorist_all_pithoi<-data.frame(aorist_all_pithoi)

@@

aorist_FOR_pithoi$cemetery<-"FC"
aorist_KNC_pithoi$cemetery<-"KNC"
aorist_all_pithoi$cemetery<-"All tombs"
aorist_pithoi<-rbind(aorist_all_pithoi,aorist_KNC_pithoi,aorist_FOR_pithoi)
aorist_pithoi$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FC"))

aorist_pithoi_plot<-ggplot(NULL)+
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 43, ymax = 45), fill=NA, colour="black",alpha = 0.4) +
  geom_text(data=rects,aes(xmid, 44,label=col),family="serif",size=3)+
  geom_line(data=aorist_pithoi,aes(decade,aoristic_sum,linetype=cemetery))+ 
       theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          plot.margin=unit(c(0.1,1,0.5,1),"cm"))+
            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
                scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

aorist_KNC <- colSums(KNC[160:209])
aorist_KNC<-data.frame(aorist_KNC)

@@

aorist_FOR <-colSums(FOR[160:209])
aorist_FOR<-data.frame(aorist_FOR)

@@

aorist_FOR$cemetery<-"FC"
aorist_KNC$cemetery<-"KNC"
aorist_all_tombs$cemetery<-"All tombs"
aorist_tombs<-rbind(aorist_all_tombs,aorist_KNC,aorist_FOR)
aorist_tombs$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FC"))

aorist_pots_plot<-ggplot(NULL)+
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 220, ymax = 230), fill=NA, colour="black",alpha = 0.4) +
  geom_text(data=rects,aes(xmid, 225,label=col),family="serif",size=3)+
  geom_line(data=aorist_tombs,aes(decade,aoristic_sum,linetype=cemetery))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          plot.margin=unit(c(0.1,1,0.1,1),"cm"))+
            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                  scale_y_continuous(expand=c(0,0),limits=c(0,230))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

ggarrange(aorist_pithoi_plot,aorist_pots_plot,nrow=2,labels=c("A","B"),align="v",common.legend=TRUE, legend="bottom")
```

And, despite differences in scale and timing, it is immediately apparent from Figure \@ref(fig:aoristbycem) that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods. In both cemeteries, the rate of cremation burial began rising around the end of the 9th century BC, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th. Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity. We might assume that these trends would be mirrored in the construction of new tombs, but this proves not to be the case.

```{r tombconstruction, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations. In both cases, tombs lacking any securely dateable pottery are excluded."}

@@

cbPalette <- c("#999999","#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=forcats::fct_rev(tomb_type_condensed)))+geom_bar()+
        theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Count of New Tombs")+scale_fill_manual(values=cbPalette,
                                                                                        labels=c("Unknown","Shaft Grave","Pit Cave",
                                                                                        "Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
             scale_y_continuous(expand=c(0,0),limits=c(0,25))+facet_wrap(~cem_condensed)+
               theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA))

new_tomb_aorist_plot<-ggplot()+geom_line(data=aorist_tomb_periods_melt,aes(period,aoristic_sum,group=cemetery,linetype=cemetery))+
  theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Aoristic Sum of New Tombs")+
               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," FC"))+ 
                 scale_y_continuous(expand=c(0,0),limits=c(0,28))+scale_x_discrete(expand=c(0,0))

ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,align="h",labels=c("A","B"), widths=c(1.6,1),common.legend=FALSE, legend="bottom")

@@

KNC_tombs_in_use<-data.frame(colSums(KNC[6:15]!=""))
colnames(KNC_tombs_in_use)<-c("no_tombs")
setDT(KNC_tombs_in_use, keep.rownames = "period")
KNC_tombs_in_use$period<-factor(KNC_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
KNC_tombs_in_use$cemetery<-"KNC"

FOR_tombs_in_use<-data.frame(colSums(FOR[6:15]!=""))
colnames(FOR_tombs_in_use)<-c("no_tombs")
setDT(FOR_tombs_in_use, keep.rownames = "period")
FOR_tombs_in_use$period<-factor(FOR_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
FOR_tombs_in_use$cemetery<-"FOR"

@@

tombs_in_use<-data.frame(colSums(tombs[6:15]!=""))
colnames(tombs_in_use)<-c("no_tombs")
setDT(tombs_in_use, keep.rownames = "period")
tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))

ggplot(tombs_in_use_cem,aes(period,no_tombs, linetype=cemetery,group=cemetery))+geom_line()+theme_classic()+
       theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Tombs in Use")+
              scale_y_continuous(expand=c(0,0),limits=c(0,45))+
              scale_x_discrete(expand=c(0,0))
```

@@

```{r tombsinusebybuilddate, fig.dim=c(7,5), dpi=600, fig.cap="Estimates of the number of tombs in use through time across both cemeteries. The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
tombs_11th<-subset(tombs,tombs$century_built=="11th")
tombs_10th<-subset(tombs,tombs$century_built=="10th")
tombs_9th<-subset(tombs,tombs$century_built=="9th")
tombs_8th<-subset(tombs,tombs$century_built=="8th")
tombs_7th<-subset(tombs,tombs$century_built=="7th")

@@

ggplot()+geom_area(data=tombs_in_use,aes(period,no_tombs,fill="All tombs",group=1),alpha=0.2)+
            geom_line(data=tomb_use,aes(period,no_tombs,linetype=tomb_century,group=tomb_century))+
              theme_bw()+theme(text=element_text(family="serif", size=12),
          plot.margin=unit(c(0.5,1,0.5,1),"cm"),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = "bottom")+
                 labs(x="Period", y="Tombs in use",linetype="Century of tomb construction",fill="")+
                  scale_fill_manual(values="#999999")+
                    scale_y_continuous(expand=c(0,0),limits=c(0,60))+
                      scale_x_discrete(expand=c(0,0))

@@

```{r potsbydateandtombdate,  fig.dim=c(6,4),out.width="0.5%", dpi=600, fig.cap="Counts of vessels from all tombs in both cemeteries, arranged, above, by the century to which they are assigned based on their ceramic period and, below, by the centuries in which the tombs they come from were built."}

@@

```{r cluster, fig.dim=c(3.4,3.9),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb. Ward's method was used and tombs that yielded no pots at all were excluded." }
